Prospective testing for the prevalence or transience of a shock effect over a set time horizon before it occurs

### 2022-03-12 prevalence_testing/postshocktest2021draft.tex
addition of statements related to weighted AIC
added new procedures for weighted AIC in the code and manuscript

% use packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{url}

@@

Suppose we are interested in comparing two models $\mc{M}_1$ and $\mc{M}_2$ using AIC for forecasting the response in the time series of interest. However, direct comparison is not feasible as the data are not observed. Instead, we can conduct such inference based on the donor pool. Let the AIC of $\mc{M}_j$ for $i$th time series be $a_i^j$ for $i = 1, \ldots, n+1$ and $j = 1, 2$. Let $\xi_i = \P(a_i^1 > a_i^2)$. Note that if the data generating process is based on $\mc{M}_1$ and is true for all the donors and time series of interest, $\xi_i$ is expected to be high. Additionally, it is reasonable to assume $\xi_i =\xi $ for $\xi \in [0,1]$ and $i = 1, \ldots, n+1$.  Notice that this setup is similar to the setup of weighted voting or voting we discussed before. So, under similar arguments, we propose the following corollary for synthetic pairwise model selection.

\begin{corollary}[Synthetic pairwise model selection]
\label{coro1}Let $(w_2, \ldots, w_{n+1})$ be weights such that $w_i \in [0,1]$ and $\sum_{i=2}^{n+1} w_i=1$. Define
\begin{align*}
  \mc{I}_n = \{i =  2,\ldots, n+1\colon 0< w_i < 1\}.
\end{align*}
Suppose that  $\mc{I}_n$  is non-empty, $|\mc{I}_n| \to \infty$ as $n \to \infty$, and $w_i b_n\leq K$ for $i\in \mc{I}_n$ and some $K>0$, where $b_n \geq 1$ and $b_n \to \infty$ as $n\to \infty$. Assume for $i \in \mc{I}_n$,  $\{a_i^1, a_i^2\}$ are pairwise independent AICs of $\mc{M}_1$ and  $\mc{M}_2$ for $i$th time series  with 
\begin{align*}
  \sum_{i\in \mc{I}_n}\P(a_i^1 > a_i^2) = \sum_{i\in \mc{I}_n} w_i \xi_i \to \xi_{1},
\end{align*}
where $\xi_i =  \P(a_i^1 > a_i^2)\in [0,1]$ for $i \in \mc{I}_n$. If $\xi  \neq 0.5$, as $n\to \infty$, 
\begin{align*}
  \E \bigg\{\left|I\left\{ \sum_{i\in \mc{I}_n} w_i I(a_i^1 > a_i^2) \geq 0.5\right\}
  - I( a_1^1 > a_1^2) \right|\bigg\}
  \to  \begin{cases}
    1- \xi_1 & \text{ if } \xi_1 > 0.5 \\
   \xi_1 & \text{ if } \xi_1 < 0.5 
  \end{cases}
\end{align*}
\end{corollary}

@@

To analyze the dataset, we propose two models. Denote the covariate vector at time $t$ for $i$th time series by $\bs{x}_{i,t}$.  The first model $\mc{M}_1$, called ``permanent'' model, is 
 \begin{align*}
 	y_{i,t} &= \eta_i + \sum_{j=1}^{q_1} \phi_{i,j} y_{i, t-j}  +\sum_{j=0}^{q_2-1} \bs{x}_{i, t-j}\bs{\theta}_{i, j + 1}  + \alpha_i  D_{i,t}+ \varepsilon_{i,t},\\
 	\alpha_i & = \mu_{\alpha} + \sum_{j=0}^{q_2-1} \bs{x}_{i, T_i^*+1-j}\bs{\delta}_{i, j + 1} + \varepsilon_{\alpha, i},
 \end{align*}
where $q_1 = q_2=2$, $D_{i,t} = I(t > T_i^*+1)$, $i = 1, \ldots, n+1$ and $t = 1, \ldots, T_i$. Note that we allow the shock effect to take effect for multiple time points instead of one time point as in  \cite{lin2021minimizing}. This simple model describes the setting where one additive shock effect is persistent over time. The second model $\mc{M}_2$, called ``dynamic'' model, is
\begin{align*}
 	y_{i,t}
 	&= \left(\eta_i + \sum_{j=1}^{q_1} \phi_{i,j} y_{i, t-j} + \sum_{j=0}^{q_2-1} \bs{x}_{i, t-j}\bs{\theta}_{i, j + 1} \right) (1-D_{i,t}) + f(\mc{F}_{i, t} \alpha_i)D_{i,t} + \varepsilon_{i,t}  \\
 	f(\mc{F}_{i,t},\alpha_i)	 & = \alpha_i +\sum_{j=1}^{q_1} \tilde{\phi}_{i, j} y_{i, t-j} + \sum_{j=0}^{q_2-1} \bs{x}_{i, t-j} \tilde{\bs{\theta}}_{i, j +1} \\
 	\alpha_i & = \mu_{\alpha} + \sum_{j=0}^{q_2-1} \bs{x}_{i, T_i^*+1-j}\bs{\delta}_{i, j + 1} + \varepsilon_{\alpha, i},
 \end{align*}
 where $q_1 = q_2=2$,  $D_{i,t} = I(t > T_i^*+1)$, $i = 1, \ldots, n+1$, and $t = 1, \ldots, T_i$. We estimate $\mc{M}_1$ and $\mc{M}_2$ using ordinary least squares (OLS) estimation. $\mc{M}_2$ adds more flexibility to $\mc{M}_1$ by allowing that the model of $y_{i,t}$ changes after the shock, as seen from the term related to $1-D_{i,t}$,  though the model still preserves the autoregressive model form. In practice, $\mc{M}_2$ should make more sense than $\mc{M}_1$ as it accommodates for more complicated cases. However, $\mc{M}_2$  requires far more degrees of freedom for estimation and this can increase variance in estimation. We used weighted AIC for model selection, which is detailed as follows. For $i$th donor in the donor pool, we can compute its AICs for $\mc{M}_1$ and $\mc{M}_2$, say $a^1_i$ and $a^2_i$, and also its synthetic weight $W_i^*$, where $\mbf{W}^* = (w_2^*, \ldots, w_{n+1}^*)$ and its computation is detailed in the next paragraph. Let $A^{1*}= \sum_{i=2}^{n+1} w_i^* a^1_i$ and $A^{2*}= \sum_{i=2}^{n+1} w_i^*a^2_i$. We will select $\mc{M}_1$ if $A^{1*} < A^{2*}$ and vice versa.

@@

\begin{table}[H]
	\caption{Results for Conoco Philips ($n=3$), $\mbf{W}^* = (0.183, 0, 0.817)$ from scaled covariates} \label{coptable}
\begin{center}
  \begin{tabular}{cccccccccr}
     $L$  & Model & weighted  AIC & $p_1$ &  $|\hat{p}-p_1|$ & Vote (Correct?) & weighted Vote (Correct?)   \\
     \hline 
     7 & $\mc{M}_1$  & $214.10$ & 1  & 0 & 0 (Yes) &0 (Yes)  \\      
     7 & $\mc{M}_2$  & 205.23 & 1 & $0$   & 0 (Yes) &0 (Yes)   \\[.2cm]
     15 & $\mc{M}_1$ & 246.67 & 1   & $0.82$ & 0 (Yes) &1 (No)    \\
     15 &  $\mc{M}_2$ & $238.24$ & 1 & 0 &0 (Yes)  &0 (Yes)  \\[.2cm]
     30 &  $\mc{M}_1$ & 303.43 & 1   & $0.82$ & 0 (Yes) &1 (No)   \\
     30 &  $\mc{M}_2$ & 293.38 & 1  & 0 &0 (Yes) & 0 (Yes)  
  \end{tabular}
\end{center}	
\end{table}

@@

In term of Table \ref{coptable}, it is important to note that $p_1$ is the $p$-value of the forecast comparison between adjusted forecast and unadjusted forecast if we have the data $\{y_{1,t} \colon 1, \ldots, T_1\}$, where $T_1 > T_1^*$. In other words, it cannot be computed without observing $y_{1, t}$ for $t > T_1^*$ and we present $p_1$ only for retrospective analysis. Also, $\hat{p}$ is the $p$-value from weighting the $p$-values of the forecast comparison  between adjusted and unadjusted forecast for donors using $\mbf{W}^*$. From Table \ref{coptable}, for all $L = 7, 15, 30$, $\mc{M}_2$ is preferred over $\mc{M}_1$. Notice that $|\hat{p}-p_1|$ is uniformly zero and $p_1$ is uniformly 1 for different $L$ and $\mc{M}_2$. The test  using $\hat{p}$ with $\mc{M}_2$ fails to reject that  the shock, which was to take effect at $t=T_1^*+1$,  is prevalent for $t > T_1^*+1$ under a significance level of 5\%. Nevertheless, the results appear to be a little different for $\mc{M}_1$. For $L=7$, $|\hat{p}-p_1|=0$ for $\mc{M}_1$ whereas $|\hat{p}-p_1|=0.82$ for $L = 15, 30$. Based on the performance of $\mc{M}_2$, it is possible that as the time window $L$ increases, the signal from changing parameters covariates is ``misunderstood'' by $\mc{M}_1$ as shock signal.

@@

In table 2, we observe that weighted AIC prefers $\mc{M}_2$, and thus we turn our interest to that model.  Since $p_{1}$ is far above the rejection threshold, we lack the evidence to reject the null of a transient shock in the time series under study, the month-on-month change in log employment in the span 2019-2022.  The weighted p-value, $\hat p $, misses the ground truth p-value by .146 in absolute value, yet renders the correct decision.

@@

@@

\begin{table}[H]
	\caption{Results for Month-on-Month Change in Log Nonfarm Payrolls (Seasonally Adjusted) ($n=3$), $\mbf{W}^* = (0, 0.278, 0.722)$ from scaled covariates} \label{uemtable}
\begin{center}
  \begin{tabular}{ccccccccr}
      Model & weighted  AIC & $p_1$ &  $|\hat{p}-p_1|$ & Vote (Correct?) & weighted Vote (Correct?)   \\
     \hline 
     $\mc{M}_1$  & $-687.836$ & 0  & 1 & 1 (No) &1 (No)  \\      
      $\mc{M}_2$ ($q_1=q_2=1$)  & $-688.428$ & 0.470 & $0.146$   & 0 (Yes) &0 (Yes)   \\[.2cm]
  \end{tabular}
\end{center}	
\end{table}

>>>>

% use packages
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amscd}
\usepackage{amssymb}
\usepackage{natbib}
\usepackage{url}
\usepackage{multirow}

@@

Suppose we are interested in comparing two models $\mc{M}_1$ and $\mc{M}_2$ using AIC for forecasting the response in the time series of interest. However, direct comparison is not feasible as the data are not observed. Instead, we can conduct such inference based on the donor pool. Let the AIC of $\mc{M}_j$ for $i$th time series be $a_i^j$ for $i = 1, \ldots, n+1$ and $j = 1, 2$. Let $\xi_i = \P(a_i^1 < a_i^2)$. Note that if the data generating process is based on $\mc{M}_1$ and is true for all the donors and time series of interest, $\xi_i$ is expected to be high. Additionally, it is reasonable to assume $\xi_i =\xi $ for $\xi \in [0,1]$ and $i = 1, \ldots, n+1$.  Notice that this setup is similar to the setup of weighted voting or voting we discussed before. So, under similar arguments, we propose the following corollary for synthetic pairwise model selection.

\begin{corollary}[Synthetic pairwise model selection]
\label{coro2}Let $(w_2, \ldots, w_{n+1})$ be weights such that $w_i \in [0,1]$ and $\sum_{i=2}^{n+1} w_i=1$. Define
\begin{align*}
  \mc{I}_n = \{i =  2,\ldots, n+1\colon 0< w_i < 1\}.
\end{align*}
Suppose that  $\mc{I}_n$  is non-empty, $|\mc{I}_n| \to \infty$ as $n \to \infty$, and $w_i b_n\leq K$ for $i\in \mc{I}_n$ and some $K>0$, where $b_n \geq 1$ and $b_n \to \infty$ as $n\to \infty$. Assume for $i \in \mc{I}_n$,  $\{a_i^1, a_i^2\}$ are AICs of $\mc{M}_1$ and  $\mc{M}_2$ for $i$th time series, which are pairwise independent of $\{a_j^1, a_j^2\}$ for $j \neq i$ and that
\begin{align*}
  \sum_{i\in \mc{I}_n}\P(a_i^1 < a_i^2) = \sum_{i\in \mc{I}_n} w_i \xi_i \to \xi_{1},
\end{align*}
where $\xi_i =  \P(a_i^1 < a_i^2)\in [0,1]$ for $i \in \mc{I}_n$. If $\xi  \neq 0.5$, as $n\to \infty$, 
\begin{align*}
  \E \bigg\{\left|I\left\{ \sum_{i\in \mc{I}_n} w_i I(a_i^1 < a_i^2) \geq 0.5\right\}
  - I( a_1^1 < a_1^2) \right|\bigg\}
  \to  \begin{cases}
    1- \xi_1 & \text{ if } \xi_1 > 0.5 \\
   \xi_1 & \text{ if } \xi_1 < 0.5 
  \end{cases}
\end{align*}
\end{corollary}

Note that Corollary \ref{coro2} gives an asymptotic upper bound for the misclassification rate of synthetic pairwise model selection, which is small if $\xi_1$ is large.

@@

To analyze the dataset, we propose two models. Denote the covariate vector at time $t$ for $i$th time series by $\bs{x}_{i,t}$.  The first model $\mc{M}_1$, called ``permanent'' model, is 
 \begin{align*}
 	y_{i,t} &= \eta_i + \sum_{j=1}^{q_1} \phi_{i,j} y_{i, t-j}  +\sum_{j=0}^{q_2-1} \bs{x}_{i, t-j}\bs{\theta}_{i, j + 1}  + \alpha_i  D_{i,t}+ \varepsilon_{i,t},\\
 	\alpha_i & = \mu_{\alpha} + \sum_{j=0}^{q_2-1} \bs{x}_{i, T_i^*+1-j}\bs{\delta}_{i, j + 1} + \varepsilon_{\alpha, i},
 \end{align*}
where $q_1 = q_2=2$, $D_{i,t} = I(t > T_i^*+1)$, $i = 1, \ldots, n+1$ and $t = 1, \ldots, T_i$. Note that we allow the shock effect to take effect for multiple time points instead of one time point as in  \cite{lin2021minimizing}. This simple model describes the setting where one additive shock effect is persistent over time. The second model $\mc{M}_2$, called ``dynamic'' model, is
\begin{align*}
 	y_{i,t}
 	&= \left(\eta_i + \sum_{j=1}^{q_1} \phi_{i,j} y_{i, t-j} + \sum_{j=0}^{q_2-1} \bs{x}_{i, t-j}\bs{\theta}_{i, j + 1} \right) (1-D_{i,t}) + f(\mc{F}_{i, t} \alpha_i)D_{i,t} + \varepsilon_{i,t}  \\
 	f(\mc{F}_{i,t},\alpha_i)	 & = \alpha_i +\sum_{j=1}^{q_1} \tilde{\phi}_{i, j} y_{i, t-j} + \sum_{j=0}^{q_2-1} \bs{x}_{i, t-j} \tilde{\bs{\theta}}_{i, j +1} \\
 	\alpha_i & = \mu_{\alpha} + \sum_{j=0}^{q_2-1} \bs{x}_{i, T_i^*+1-j}\bs{\delta}_{i, j + 1} + \varepsilon_{\alpha, i},
 \end{align*}
 where $q_1 = q_2=2$,  $D_{i,t} = I(t > T_i^*+1)$, $i = 1, \ldots, n+1$, and $t = 1, \ldots, T_i$. We estimate $\mc{M}_1$ and $\mc{M}_2$ using ordinary least squares (OLS) estimation. $\mc{M}_2$ adds more flexibility to $\mc{M}_1$ by allowing that the model of $y_{i,t}$ changes after the shock, as seen from the term related to $1-D_{i,t}$,  though the model still preserves the autoregressive model form. In practice, $\mc{M}_2$ should make more sense than $\mc{M}_1$ as it accommodates for more complicated cases. However, $\mc{M}_2$  requires far more degrees of freedom for estimation and this can increase variance in estimation. For this purpose, the synthetic pairwise model selection proposed in the Corollary \ref{coro2} can be used to select between $\mc{M}_1$ and $\mc{M}_2$.

@@

\begin{table}[H]
	\caption{Results for Conoco Philips ($n=3$), $\mbf{W}^* = (0.183, 0, 0.817)$ from scaled covariates} \label{coptable}
\begin{center}
  \begin{tabular}{cccccccccr}
     $L$  & Model & $\sum_{i=2}^4 w_i^* I(a_i^1 > a_i^2)$ & $p_1$ &  $|\hat{p}-p_1|$ & Vote (Correct?) & weighted Vote (Correct?)   \\
     \hline 
     7 & $\mc{M}_1$  & \multirow{2}{*}{$0.817$} & 1  & 0 & 0 (Yes) &0 (Yes)  \\      
     7 & $\mc{M}_2$  &  & 1 & $0$   & 0 (Yes) &0 (Yes)   \\[.2cm]
     15 & $\mc{M}_1$ & \multirow{2}{*}{$0.817$}  & 1   & $0.82$ & 0 (Yes) &1 (No)    \\
     15 &  $\mc{M}_2$ & & 1 & 0 &0 (Yes)  &0 (Yes)  \\[.2cm]
     30 &  $\mc{M}_1$ & \multirow{2}{*}{$1$}  & 1   & $0.82$ & 0 (Yes) &1 (No)   \\
     30 &  $\mc{M}_2$ &  & 1  & 0 &0 (Yes) & 0 (Yes)  
  \end{tabular}
\end{center}	
\end{table}

@@

In term of Table \ref{coptable}, it is important to note that $p_1$ is the $p$-value of the forecast comparison between adjusted forecast and unadjusted forecast if we have the data $\{y_{1,t} \colon 1, \ldots, T_1\}$, where $T_1 > T_1^*$. In other words, it cannot be computed without observing $y_{1, t}$ for $t > T_1^*$ and we present $p_1$ only for retrospective analysis. Also, $\hat{p}$ is the $p$-value from weighting the $p$-values of the forecast comparison  between adjusted and unadjusted forecast for donors using $\mbf{W}^*$.

Let $a_i^1$ and $a_i^2$ denote the AICs of $\mc{M}_1$ and $\mc{M}_2$ for $i$th time series, respectively. Using the synthetic pairwise model selection procedures specified in Corollary \ref{coro2}, as  $\sum_{i=2}^4 w_i^* I(a_i^1 > a_i^2)>0.5$ for $L = 7, 15, 30$, we conclude that $\mc{M}_2$ is preferred over $\mc{M}_1$.  Notice that $|\hat{p}-p_1|$ is uniformly zero and $p_1$ is uniformly 1 for different $L$ and $\mc{M}_2$. The test  using $\hat{p}$ with $\mc{M}_2$ fails to reject that  the shock, which was to take effect at $t=T_1^*+1$,  is prevalent for $t > T_1^*+1$ under a significance level of 5\%. Nevertheless, the results appear to be a little different for $\mc{M}_1$. For $L=7$, $|\hat{p}-p_1|=0$ for $\mc{M}_1$ whereas $|\hat{p}-p_1|=0.82$ for $L = 15, 30$. Based on the performance of $\mc{M}_2$, it is possible that as the time window $L$ increases, the signal from changing parameters covariates is ``misunderstood'' by $\mc{M}_1$ as shock signal.

@@

@@

In Table 2, using the synthetic pairwise model selection procedures specified in Corollary \ref{coro2}, because $\sum_{i=2}^4 w_i^* I(a_i^1 > a_i^2)>0.5$,  we conclude that  $\mc{M}_2$ is preferred. Thus, we turn our interest to that model.  Since $p_{1}$ is far above the rejection threshold, we lack the evidence to reject the null of a transient shock in the time series under study, the month-on-month change in log employment in the span 2019-2022.  The weighted p-value, $\hat p $, misses the ground truth p-value by .146 in absolute value, yet renders the correct decision.

@@

\begin{table}[H]
	\caption{Results for Month-on-Month Change in Log Nonfarm Payrolls (Seasonally Adjusted) ($n=3$), $\mbf{W}^* = (0, 0.278, 0.722)$ from scaled covariates} \label{uemtable}
\begin{center}
  \begin{tabular}{ccccccccr}
      Model & $\sum_{i=2}^4 w_i^* I(a_i^1 > a_i^2)$  & $p_1$ &  $|\hat{p}-p_1|$ & Vote (Correct?) & weighted Vote (Correct?)   \\
     \hline 
     $\mc{M}_1$  & \multirow{2}{*}{0.722}& 0  & 1 & 1 (No) &1 (No)  \\      
      $\mc{M}_2$ ($q_1=q_2=1$)  & & 0.470 & $0.146$   & 0 (Yes) &0 (Yes)   \\[.2cm]
  \end{tabular}
\end{center}	
\end{table}
